\thetitle

### 2021-05-07 writing/discussion.tex
Touch up previous edits
Add reference for Mill's homo economicus

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently, the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating is a
mistake. To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

>>>>

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently (e.g.~\cite{henrich2001search}), the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

### 2021-05-07 writing/why-rational.tex
Note that RP re-runs program in the same way
Add reference for Mill's homo economicus

This paper dubs Lazarek et. al's idea {\em the rational programmer\/}. Like Stuart
Mill's {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task by exploiting the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Specifically,
the rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch by adding type
annotations to the untyped parts of the program identified in the error.
If the type checker rejects an annotation that is considered correct, the
rational programmer has found the source of the problem.
Otherwise, the rational programmer concludes that the just-annotated parts are not the problem
and re-runs the program;
the Wadler-Findler slogan says the new version of the program must now blame a different location for the problem, which the rational programmer can check in the same fashion.
Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in a human-facing study.

@@

Assuming the rational programmer trusts the type of {\tt na-client}, the next
step is to inspect the {\tt layer} module and to equip it with type
annotations. The result is the \typecolor\ module in the middle, and {\tt
main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. As
predicted by the theory, running the modified program yields a different error message:
\begin{verbatim}
    neg-abs: broke its own contract
      promised: (<=/c 0)
      produced: 10
      in: (-> any/c (<=/c 0))
      contract from: (interface for neg-abs)
      blaming: (interface for neg-abs)
       (assuming the contract is correct)
\end{verbatim}
The last step of the rational programmer is to assign types to the {\tt server}
module and to re-direct the import of {\tt layer-typed} to {\tt (submod ".."
server-typed)}. At this point the type checker objects to the conjectured type
of {\tt neg-abs}---and the source of the impedance mismatch is found. How to fix
it is a separate question.

>>>>

This paper dubs Lazarek et. al's idea {\em the rational programmer\/}.
Like~\citet{mill1874essays}'s {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task by exploiting the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Specifically,
the rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch by adding type
annotations to the untyped parts of the program identified in the error.
If the type checker rejects an annotation that is considered correct, the
rational programmer has found the source of the problem.
Otherwise, the rational programmer concludes that the just-annotated parts are not the problem
and re-runs the program;
the Wadler-Findler slogan says the new version of the program must now blame a different location for the problem, which the rational programmer can check in the same fashion.
Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in a human-facing study.

@@

Assuming the rational programmer trusts the type of {\tt na-client}, the next
step is to inspect the {\tt layer} module and to equip it with type
annotations. The result is the \typecolor\ module in the middle, and {\tt
main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. As
predicted by the theory, running the modified program (in the same way as before) yields a different error message:
\begin{verbatim}
    neg-abs: broke its own contract
      promised: (<=/c 0)
      produced: 10
      in: (-> any/c (<=/c 0))
      contract from: (interface for neg-abs)
      blaming: (interface for neg-abs)
       (assuming the contract is correct)
\end{verbatim}
The last step of the rational programmer is to assign types to the {\tt server}
module and to re-direct the import of {\tt layer-typed} to {\tt (submod ".."
server-typed)}. At this point the type checker objects to the conjectured type
of {\tt neg-abs}---and the source of the impedance mismatch is found. How to fix
it is a separate question.
